The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2019-10-12 categorical-zero-one.tex
started adding more text

\documentclass{article}
\usepackage{eigilscmds}
\usepackage{tikzit}
\input{/home/eigil/Downloads/tikzit/styles/foo.tikzstyles}
\usepackage{todonotes}
\author{Eigil Fjeldgren Rischel}
\title{A Kolmogorov zero to one law for Markov Categories}
\date{}
\renewcommand{\sf}{\mathsf}
\DeclareMathOperator{\del}{del}
\begin{document}
\maketitle
\section{Introduction}
This is a treatment of Kolmogorov's zero to one law in the setting of Markov categories.
Throughout, $J$ is a possibly infinite set and $F \subset J$ is a finite subset.

\section{Definitions and preliminaries}
Suppose $\cC$ is a Markov category.
To make sense of infinite families of random variables,
we will want to have a notion of infinite tensor product.
Of course, there is no reason such a thing should exist - for instance, it would be very strange for $\sf{FinStoch}$ to have infinite tensor products.
But in cases where the subcategory $\cC_{det}$ of deterministic maps has infinite products as well as finite ones, it seems natural to ask whether this is compatible with the structure of $\cC$ in a suitable sense.
We introduce the following notion
\begin{definition}
    Let $\{X_i\}_{i \in J}$ be a collection of objects in $\cC$.
    Then we may form the diagram \[F \mapsto \bigotimes_{i \in F} X_i\]
    indexed over the poset of finite subsets $F \subset J$, ordered by reverse inclusion, where the structure maps are induced by $\del_{X_i}: X_i \to I$ and the coherence isomorphisms.

    We define $\bigotimes_{i \in J}X_i$ to be the limit of this (cofiltered) diagram in $\cC$, if it exists.
\end{definition}
\begin{remark}
    If $J$ is finite, this coincides with the normal tensor product indexed over $J$.
\end{remark}
\begin{remark}
    It is easily verified that 
    \[\bigotimes_{i \in J} X_i \tensor \bigotimes_{i' \in J'} X_{i'} \cong \bigotimes_{i \in J \sqcup J'} X_i\]
    i.e tensor products of infinite tensor products can be written as single infinite tensor products in the obvious way.

    Probably the similar theorem about infinite tensor products of infinite tensor products is true as well, although I have not verified it. 
\end{remark}
\begin{remark}
    Supposing $\bigotimes_{i\in J}X_i$ exists, the structure morphisms $\pi_j: \bigotimes_{i \in J} X_i \to X_j$ are deterministic and exhibit it as the product in $\cC_{det}$ of $\{X_i\}$.
\end{remark}
\begin{remark}
    If we take as input certain maps $I \to X_i$ as well, so that we have an infinite system of probability distributions, we can also try to define the infinite tensor product as a colimit, which will then be a space with a probability distribution, corresponding to an infinite system of random variables.
    It would also be possible to formulate the zero-to-one law in this setting, but I have not explored this.
\end{remark}

To actually use this idea to reproduce the usual Kolmogorov zero-one law, we need to do a bit of work:
\begin{example}
Let $\{(X_i,\Sigma_i)\}_{i\in J}$ be a set of measurable spaces in $\sf{Stoch}$.
Then consider their product $X = \prod_i X_i$ with the $\sigma$-algebra generated by the projections $\pi_i: X \to X_i$, which we denote $\Sigma$.
Let us write in general $X_F = \prod_{i\in F}X_F$, $\Sigma_F$ for the corresponding $\sigma$-algebra, and $\pi_F: X \to X_F$ for the projection.

@@

@@

For instance, maybe there is a Markov category with objects the profinite sets, which can be obtained from $\sf{FinStoch}$ by some formal construction (maybe even simply as the pro-category), which admits infinite tensor products, and where one can profitably interpret this theorem.
\end{document}

>>>>

\documentclass[11pt]{article}
\usepackage{eigilscmds}
\usepackage{tikzit}
\usepackage{geometry}
\input{foo.tikzstyles}

% draft packages
\usepackage{todonotes}
\usepackage{showkeys}

\author{Tobias Fritz and Eigil Fjeldgren Rischel}
\title{The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability}
\date{\today}

\renewcommand{\sf}{\mathsf}
\DeclareMathOperator{\del}{del}
\begin{document}
\maketitle

\begin{abstract}
	We state and prove the zero--one laws of Kolmogorov and Hewitt--Savage within the setting of \emph{Markov categories}, a category-theoretic approach to the foundations of probability and statistics. This gives general versions of these results which can be instantiated not only in measure-theoretic probability, where they specialize to the standard ones, but also in other contexts such as categories of closed relations between topological spaces, 
	\improve{T: this abstract states what we can hopefully achieve. To be adjusted}
\end{abstract}

\tableofcontents

\section{Introduction}

This is a treatment of the zero--one laws of Kolmogorov and of Hewitt--Savage in the setting of Markov categories.

\paragraph*{Notation.}

Throughout, $\cC$ is a Markov category in the sense of \cite[Definition~2.1]{markov_cats} (see also the closely related earlier work of Golubtsov~\cite{golubtsov} and Cho and Jacobs~\cite{cho_jacobs}). $J$ is a possibly infinite set, used for indexing products of objects, and $F \subset J$ ranges over finite subsets.

\section{Background on Markov categories}

\todo[inline]{to be written}

\section{Infinite tensor products}

Suppose $\cC$ is a Markov category.
To make sense of infinite families of random variables within $\cC$,
we will want to have a notion of infinite tensor product in $\cC$. By Kolmogorov's extension theorem, the joint distribution of infinitely many random variables is uniquely determined by the family of distributions of all finite subsets of these variables. This motivates a general definition of the infinite tensor product of a family of objects $(X_i)_{i \in J}$, based on the observation that if $F \subseteq F' \subseteq J$ are two finite subset, then the fact that $\cC$ is semicartesian monoidal gives us morphisms
\[
	\bigotimes_{i \in F'} X_i \longrightarrow \bigotimes_{i \in F} X_i
\]
to be interpreted as \emph{marginalization}. Via these maps, the finite tensor products $\bigotimes_{i \in F} X_i$ make up a cofiltered diagram in the form of a functor from the poset of finite subsets of $J$, ordered by reverse inclusion, to $\cC$.

\begin{definition}
    Let $(X_i)_{i \in J}$ be a family of objects in $\cC$.
    Then the \emph{infinite tensor product} $\bigotimes_{i \in J} X_i$ is the limit of the diagram \[F \longmapsto \bigotimes_{i \in F} X_i,\]
    indexed by the poset of finite subsets $F \subset J$ ordered by reverse inclusion, if this limit exists.
\end{definition}

Similar definitions of infinite tensor products of algebraic structures as \emph{filtered colimits} of finite tensor products are well-known in the literature, e.g.~in the case of C*-algebras~\cite[p.~315]{blackadar}. Intuitively, our definition matches up with these under the categorical duality of algebra and geometry, where our definition is of a geometrical flavour.

Of course, in a general Markov category the above limit need not exist, in which case the infinite tensor product does not exist either (\Cref{infprods_finstoch}).
But in cases where the subcategory $\cC_{\det}$ of deterministic maps has infinite products as well as finite ones, it seems natural to ask whether this is compatible with the structure of $\cC$ in a suitable sense.
We introduce the following notion

\begin{remark}
	If $J$ is finite, the infinite tensor product coincides with the standard tensor product $\bigotimes_{i \in J} X_i$.
\end{remark}

\begin{remark}
    It is easily verified that 
    \[\bigotimes_{i \in J} X_i \tensor \bigotimes_{i' \in J'} X_{i'} \cong \bigotimes_{i \in J \sqcup J'} X_i\]
    i.e tensor products of infinite tensor products can be written as single infinite tensor products in the obvious way.

    Probably the similar theorem about infinite tensor products of infinite tensor products is true as well, although I have not verified it. 
\end{remark}

\begin{remark}
    Supposing $\bigotimes_{i\in J}X_i$ exists, the structure morphisms $\pi_j: \bigotimes_{i \in J} X_i \to X_j$ are deterministic and exhibit it as the product in $\cC_{\det}$ of $\{X_i\}$.
\end{remark}

\begin{remark}
    If we take as input certain maps $I \to X_i$ as well, so that we have an infinite system of probability distributions, we can also try to define the infinite tensor product as a colimit, which will then be a space with a probability distribution, corresponding to an infinite system of random variables.
    It would also be possible to formulate the zero-to-one law in this setting, but I have not explored this.
\end{remark}

To actually use this idea to reproduce the usual Kolmogorov zero-one law, we need to do a bit of work:

\begin{example}
Let $\{(X_i,\Sigma_i)\}_{i\in J}$ be a set of measurable spaces in $\sf{Stoch}$.
Then consider their product $X = \prod_i X_i$ with the $\sigma$-algebra generated by the projections $\pi_i: X \to X_i$, which we denote $\Sigma$.
Let us write in general $X_F = \prod_{i\in F}X_F$, $\Sigma_F$ for the corresponding $\sigma$-algebra, and $\pi_F: X \to X_F$ for the projection.

@@

\begin{example}
	\label{infprods_finstoch}
	$\sf{FinStoch}$ does not have infinite tensor products.
\end{example}

@@

For instance, maybe there is a Markov category with objects the profinite sets, which can be obtained from $\sf{FinStoch}$ by some formal construction (maybe even simply as the pro-category), which admits infinite tensor products, and where one can profitably interpret this theorem.

\bibliographystyle{plain}
\bibliography{categorical-zero-one}

\end{document}
